A comparison of two second order traffic flow models

### 2015-03-13 FirstDraftMaterials/Draft.tex
Updated intro and PW section

\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{subfigure}
\usepackage{enumerate}
\setlength\parindent{0pt}

@@

\begin{document}

% Introduction
\section{Introduction}
When a model is called ``higher order", we usually associate it with being more accurate than a first order model.  This is not always the case.  Adding complexity to a model does not necessarily improve it, as is demonstrated in the class of higher order traffic models developed with the intention of improving the first order traffic model proposed by Lighthill and Whitham (1955) and Richards (1956).  As Daganzo observes in his paper, \\

``A new theory (e.g. relativity) is deemed successful if it explains previously unexplained phenomena, plus everything else that was correctly explained with the theory it intends to replace (e.g. Newtonian physics).  In the case of traffic flow, the new theories (high-order fluid models) fail to explain the behavior of traffic at the end of a queue (unlike the simpler, older LWR model; which explains it perfectly)." \\

In this paper we examine the second order traffic model developed by Payne (1971) and Whitham (1974) which was meant to improve how the LWR model explains traffic behavior near shocks.  Their model takes its form from equations used to describe compressible fluid dynamics.  Unfortunately, cars and fluid particles have some fundamental differences which make the model actually perform worse than the LWR model in certain cases.  For example, by smoothing out discontinuities, the PW model predicts negative velocities, which is clearly a problem.  \\

In addition to considering the PW model, we also take a look a more successful model proposed by Aw and Rascle (2000).  These authors challenge some of the assumptions of the PW model and suggest the use of a convective derivative to fix the problem of negative velocities.  In order to verify that their model is valid, they test their new model against criteria they claim any reasonable model will satisfy.

% LWR Model
\section{LWR Model}
We can start by considering a first order (one equation) model of traffic flow on a one-way highway without any exits, entrances, or passing.  In this case, we can expect that the number of cars will be conserved, so we can use the equation for conservation of mass (with density $\rho$ and velocity $v$):

\begin{enumerate}[(a)]
\item Integral form of equation: $\displaystyle \frac{d}{dt} \int_{x_1}^{x_2} \rho \, dt = -\Big[ f( \rho ) \Big]_{x_1}^{x_2}$
\item Differential form of equation: $\rho_t + f( \rho )_x = 0$
\begin{enumerate}[i]
\item Light traffic: $\rho_t + (\rho v)_x = 0$
\item Congested traffic: $\rho_t + \Big( \rho V(\rho) \Big)_x = 0$ (LWR model)
\end{enumerate}
\end{enumerate}

The LWR behaves well macroscopically, but it does have some flaws.  For example, it assumes cars change their velocity instantaneously as they travel through shocks.  One way to eliminate shocks would be to add a diffusion term: $\rho_t + \Big( \rho V(\rho) \Big)_x = \mu \rho_{xx}$.  This term is meant to account for the drivers awareness of the road ahead.  However, this model can be shown to predict negative velocities in some instances, which is obviously wrong.

@@

% PW Model: Theory
\subsection{Theory}
Another approach to improving the LWR model would be to use a second order (two equation) model. Since cars can be thought of as a compressible gas, we can use equations for conservation of mass (above) and conservation of momentum, $\rho v$, used to model compressible gas flow:

\begin{enumerate}[(a)]
\item Integral form of equation: $\displaystyle \frac{d}{dt} \int_{x_1}^{x_2} \rho v \, dt = -\Big[ f( \rho v ) \Big]_{x_1}^{x_2}$
\item Differential form of equation: $(\rho v)_t + f( \rho v )_x = (\rho v)_t + \Big( \rho v^2 + \rho \Big)_x = 0$
\end{enumerate}

We can rewrite the second equation in terms of velocity instead of momentum: $v_t + v v_x + \dfrac{p'(\rho)}{\rho} \rho_x = 0$. \\

The PW model is a modification of these equations for conservation of mass and conservation of momentum in gas dynamics, with added relaxation and diffusion terms:

@@

At first glance, this seems like a reasonable model.  However, there a couple of fundamental differences between gas particles and cars that could be problematic.  For example, as pointed out by Daganzo, ``a fluid particle responds to stimuli from the front and from behind, but a car is an anisotropic particle that mostly responds to frontal stimuli."  If we consider the eigenvalues of the Jacobian matrix of the linearized system, we see that one of the eigenvalues, $\lambda_2 = v + \sqrt{p'(\rho)}$, represents information traveling faster than the velocity of cars in the system.  This implies that traffic conditions are partially determined by what is going on behind.

>>>>

\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{subfigure}
\usepackage{enumerate}

@@

\title{A comparison of two second order traffic flow models}
\author{Kelsey Maass and Brisa Davis}
\begin{document}
\maketitle

% Introduction
\section{Introduction}
When a model is called ``higher order", we usually associate it with being more accurate than a first order model.  However, this is not always the case.  Adding complexity to a model does not necessarily improve it, as is demonstrated in the class of higher order traffic flow models developed with the intention of improving the first order traffic flow model proposed by Lighthill, Whitham and Richards.  As Carlos F. Daganzo observes in his paper \cite{Daganzo1995}, ``A new theory ... is deemed successful if it explains previously unexplained phenomena, plus everything else that was correctly explained with the theory it intends to replace."  Unfortunately, some of these higher order models are not successful because they actually behave worse than the original.  \\

In this paper we first examine the second order traffic model developed by Payne and Whitham which was meant to improve how the LWR model explains traffic behavior near shocks.  Viewing traffic as a compressible fluid, the model borrows equations used to describe compressible fluid dynamics.  Unfortunately, due to the anisotropic nature of cars, the model actually performs worse than the LWR model in certain cases.  For example, by smoothing out discontinuities, the PW model sometimes predicts negative velocities.  \\

In addition to considering the PW model, we also take at a look a more successful model proposed by Aw and Rascle \cite{AwRascle2000}.  These authors challenge some of the assumptions of the PW model and suggest the use of a convective derivative to fix the problem of negative velocities.  In order to verify that their model is valid, they test their new model against criteria they claim any reasonable model will satisfy.

% LWR Model
\section{LWR Model}
We start by considering a first order (one equation) model of traffic flow on a one-way road without any exits, entrances, or passing.  In this case, we can expect that the number of cars will be conserved, so we can use the equation for conservation of mass (with density $\rho$ and velocity $v$):

\begin{enumerate}[(a)]
\item Integral form of equation: $\displaystyle \frac{d}{dt} \int_{x_1}^{x_2} \rho \, dt = -\Big[ f( \rho ) \Big]_{x_1}^{x_2}$
\item Differential form of equation: $\rho_t + f( \rho )_x = \rho_t + \Big( \rho V(\rho) \Big)_x = 0$
\end{enumerate}

The LWR model uses the flux function $f(\rho) = \rho V(\rho)$, where $V(\rho)$ represents the preferred, or equilibrium, velocity, a given nonincreasing function of $\rho$, nonnegative for $\rho$ between 0 and $\rho_m$, the ``jam" density.  The LWR model behaves well macroscopically, predicting piece-wise smooth density with transitions between regions approximated by shocks.  However, it assumes that cars change their velocities instantaneously as they travel through shocks.  One way to eliminate shocks would be to add a diffusion term: $\rho_t + \Big(\rho V(\rho))_x = \mu \rho_{xx}$.  This additional term is meant to account for the driver's awareness of the road ahead and behind.  However, this model can be shown to predict negative velocities in some instances.

@@

% PW Model: Theory
\subsection{Theory}
Another approach to improving the LWR model would be to use a second order (two equation) model.  Since cars can be thought of as a compressible fluid, we can use equations for conservation of mass (above) and conservation of momentum, $\rho v$, used to model compressible fluid flow:

\begin{enumerate}[(a)]
\item Integral form of equation: $\displaystyle \frac{d}{dt} \int_{x_1}^{x_2} \rho v \, dt = -\Big[ f( \rho v ) \Big]_{x_1}^{x_2}$
\item Differential form of equation: $(\rho v)_t + f( \rho v )_x = (\rho v)_t + \Big( \rho v^2 + p(\rho) \Big)_x = 0$
\end{enumerate}

Here the flux function contains a contribution from the momentum, $(\rho v)v = \rho v^2$, with an additional pressure term.  Since traffic doesn't really have pressure, this term is thought of as an ``anticipation factor", describing how a driver reacts to variations in density with respect to space. Rewriting the system in terms of velocity and adding relaxation and viscosity terms, we get the full PW model:

@@

The first term on the right hand side is a relaxation term meant to mimic the driver's tendency to drive at the preferred velocity, where $\tau$ represents the driver's reaction time.  The second term, like the diffusion term above, represents the driver's awareness of nearby conditions.  Both $\tau$ and $\mu$ are nonnegative constants.

At first glance, this seems like a reasonable model.  One fundamental difference between cars and fluids causes this model to predict nonphysical situations.  As pointed out by Daganzo, 	``a fluid particle responds to stimuli from the front and from behind, but a car is an anisotropic particle that mostly responds to frontal stimuli" \cite{Daganzo1995}.  If we consider the eigenvalues of the linearized system,

\[ \begin{pmatrix} \rho \\[1ex] v \end{pmatrix}_t + \begin{pmatrix} v & \rho \\[1ex] \dfrac{p'(\rho)}{\rho} & v \end{pmatrix} \begin{pmatrix} \rho \\[1ex] v \end{pmatrix}_x = \begin{pmatrix} 0 \\[1ex] 0 \end{pmatrix}, \]

we see that the largest eigenvalue, $\lambda_2 = v + \sqrt{p'(\rho})$, represents information traveling faster than the velocity of cars in the system.  This implies that traffic conditions are partially determined by what is going on behind the drivers.  This contradicts the anisotropic nature of cars.
